% The CTC saga: C2 comparison
% Stelios Sfakianakis
% Date: 2013/03/15


```{r setup, include=FALSE, cache=FALSE}
opts_chunk$set(tidy.opts=list(keep.blank.line=FALSE, width.cutoff=60), dpi=200, prompt=TRUE, comment=NA, warning=FALSE, upload.fun=imgur_upload)
```

This is the comparison between peripheral blood of cancer patients and normal individuals, which identifies genes expressed in cancer blood. We are using the following datasets:

- [GSE27562](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE27562): "In total, we collected blood from **57** women with a diagnosis of breast cancer and **37** with a benign diagnosis"  (PMID: [21781289](http://www.ncbi.nlm.nih.gov/pubmed/21781289)) Platform: Affymetrix 

- [GSE16443](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE16443): "Blood samples were collected from **121** females referred for diagnostic mammography following an initial suspicious screening mammogram. Diagnostic work-up revealed that **67** of these women had breast cancer while **54** had no malignant disease. Additionally, 9 samples from 6 healthy female controls (three pregnant women, one breast-feeding woman and two healthy controls at different timepoints in their menstrual cycle) were included." (PMID: [20078854](http://www.ncbi.nlm.nih.gov/pubmed/20078854)). Platform: Applied Biosystems (ABI) 


Loading the data and performing the merge:
```{r load, cache=TRUE, message=FALSE}
require(inSilicoMerging)
require(siggenes)

load("data/gse27562_c2.rda")
load("data/gse16443.rda")
datasets = list(GSE27562=gse27562, GSE16443=gse16443)
mgse_COMBAT <- merge(datasets, method="COMBAT");
```


We select the samples to be used for this comparison:
```{r samplesel, cache=TRUE,  message=FALSE}
ind <- gse27562$characteristics_ch1=='phenotype: Normal' | gse27562$characteristics_ch1=='phenotype: Malignant' | gse27562$characteristics_ch1=='phenotype: Pre-Surgery (aka Malignant)'
gse27562_samples = sampleNames(gse27562[,ind])
gse16443_samples = sampleNames(gse16443)

samples = c(gse16443_samples, gse27562_samples)
c2.data = mgse_COMBAT[,samples]
```

The distribution of the samples is shown in Table 1.
```{r xtable, echo=FALSE, results="asis"}
require(xtable)
print(xtable(table(c2.data$Disease, c2.data$Study), caption="Sample distribution", label="fig:dist"))
```

```{r plot, echo=TRUE, cache=TRUE, fig.cap="MDS plot of the merged samples", dev="pdf"}
plotMDS(c2.data, "Study", "Disease")
```

```{r diff, cache=TRUE}
cl = ifelse(c2.data$Disease == 'Control', 0, 1)
sam.out <- sam(exprs(c2.data), cl, B=500, rand=0xDEAD);
summary(sam.out);
findDelta(sam.out, fdr=0.05);
delta <- findDelta(sam.out, fdr=0.05, verbose=FALSE)[2,1]
sam.sum <- summary(sam.out, delta)

w <- which(sam.sum@mat.sig$d.value > 0)
num.genes.over <- length(w)
num.genes.over
```

So we find `r num.genes.over` genes overexpressed in cancer peripheral blood.

```{r siggenes, cache=TRUE}
siggenes.all <- list.siggenes(sam.out, delta)
siggenes.over <- list.siggenes(sam.out, delta)[w]

ee=exprs(c2.data)[sam.sum@row.sig.genes,]
dim(ee)
```

